\documentclass{article}

\usepackage{fontspec}
\usepackage{tikz}

\title{Beikaitoru:\\an OpenType Hershey revival}
\author{Matthew Skala}

\defaultfontfeatures{Path=otf/}

\setsansfont{Beikaitoru506}
\setmainfont{Beikaitoru406}

\usetikzlibrary{calc}

\newcommand{\SegRad}{0.79}
\newcommand{\FatSegment}[2]{%
($(#1)!\SegRad cm!90:(#2)$) %
  ..controls ($(#1)!1.12*\SegRad cm!120:(#2)$) %
         and ($(#1)!1.12*\SegRad cm!150:(#2)$).. %
($(#1)!-\SegRad cm!0:(#2)$) %
  ..controls ($(#1)!1.12*\SegRad cm!-150:(#2)$) %
         and ($(#1)!1.12*\SegRad cm!-120:(#2)$).. %
($(#1)!\SegRad cm!-90:(#2)$) -- %
($(#2)!\SegRad cm!90:(#1)$) %
  ..controls ($(#2)!1.12*\SegRad cm!120:(#1)$) %
         and ($(#2)!1.12*\SegRad cm!150:(#1)$).. %
($(#2)!-\SegRad cm!0:(#1)$) %
  ..controls ($(#2)!1.12*\SegRad cm!-150:(#1)$) %
         and ($(#2)!1.12*\SegRad cm!-120:(#1)$).. %
($(#2)!\SegRad cm!-90:(#1)$) -- %
cycle %
}

\begin{document}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Please note that this is a preliminary and incomplete version of the
Beikaitoru fonts and documentation.

\section{The Hershey fonts}

In 1967, Allen V.\ Hershey of the U.S.\ Naval Weapons Laboratory in
Dahlgren, Virginia published a set of fonts for use with computer-controlled
plotters.  At that time, very few digital fonts were available at all, and
even fewer were freely usable.  The Hershey fonts, being public domain
because of their origin as work produced by a U.S.\ Federal Government
employee in the course of his duties, quickly because popular.  They were
eventually distributed by the U.S.\ National Bureau of Standards (which is
sometimes incorrectly credited as the origin of the fonts), then
disseminated on Usenet and in many other ways.

The plotters for which the Hershey fonts were originally designed are now
long since obsolete.  The distinctive appearance of these fonts, which
results from the constraints of those plotters, is quite different from the
look of modern digital typography.  But because the Hershey fonts were (and
to some extent still are) so widely used, they have a nostalgic familiarity
for many present-day computer users.  They evoke early CAD systems,
low-budget GIS, and Turbo Pascal BGI drivers.  This package is a present-day
(2010s) revival of the Hershey fonts with current technology, based as
nearly as possible on the original source material and including many
special features omitted from other versions.

Beikaitoru is a parasite package of the Tsukurimashou project, whose home
page is at http://tsukurimashou.osdn.jp/.  Distribution packages of
Tsukurimashou include current development versions of Beikaitoru, but
Beikaitoru also has release packages of its own which may be more stable. 
Like Tsukurimashou, Beikaitoru (as a package, which includes significant
amounts of 21st-Century software by Matthew Skala beyond Hershey's
public-domain vector data) is distributed under the GNU General Public
License version 3 (included in the COPYING file), with the following
addition:

\begin{quotation}
As a special exception, if you create a document which uses this font, and
embed this font or unaltered portions of this font into the document, this
font does not by itself cause the resulting document to be covered by the
GNU General Public License. This exception does not however invalidate any
other reasons why the document might be covered by the GNU General Public
License. If you modify this font, you may extend this exception to your
version of the font, but you are not obligated to do so. If you do not wish
to do so, delete this exception statement from your version.
\end{quotation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{CRT plotters}

In order to understand the peculiarities of Hershey's fonts, it is
necessary to understand the technological environment in which he was
working.  The Naval Weapons Laboratory had a couple of Stromberg-Carlson
SC~4010 dot plotters, which operated by firing a beam of electrons onto a
phosphor-coated screen.  The resulting image would be recorded on
photographic film.  The beam passed through an aperture plate with a
selection of different holes that could shape it to look like different text
characters; then it was directed to a controllable location on the screen. 
So as long as you were content with the font built into the aperture plate,
you could plot characters easily and they would be high-resolution.  But
there was only one font consisting of 64 glyphs at one fixed size, and it
was sealed inside a big expensive vacuum tube.

If you wanted to plot anything else, you would have to use the machine in
"dot plotter" mode, by selecting a generic small round hole in the aperture
plate and then stamping the resulting circular dots at appropriate
coordinates to build up the image.  This was not a pixel-based system as we
usually imagine them, because the dots were always round and were a fair bit
larger than the accuracy to which they could be positioned; so the distance
between two lines might be adjustable in increments about a third the
smallest width of a line.  Because the dots were round, with careful
programming it was possible to smooth out most of the jaggies that would
otherwise appear in a grid-based image.  The dots were also fuzzy, with more
electrons near the centre than near the edges, and that also provided a
certain amount of built-in antialiasing.

Hershey's original report contained a set of fonts designed for the dot
plotters:  lists of coordinates at which to place dots so that they would
overlap to form glyphs.  Again, these are not the same thing as
pixel-based bitmap fonts, because the dot locations are expressed to finer
resolution than the sizes of the dots.  One thing that made it a non-trivial
exercise was that the time needed to plot each dot was significant when
multiplied by the number of dots per character and the number of characters
per page; so it was important to minimize the number of dots used while
making the characters look as good as possible.  The Hershey dot fonts,
although interesting, never became popular and are not included in this
version of Beikaitoru.  (Maybe in a future version...)

The next step up in technology was the SC~4020.  It's not clear to me from
his initial report whether Hershey actually had access to one these in 1967,
or just its specifications and output samples, but the fact that he created
a suite of fonts suitable for devices like it makes me think he must have
had access to some device at least sort of like the SC~4020.  It worked on
the same basic principle as the SC~4010, but it could also sweep the
electron beam smoothly from one grid point to another.  Images formed in
this vector-plotting mode would consist of fattened line segments (the set
of all points within a certain fixed radius of any point on any of a set of
line segments) where the endpoints were all on an integer grid, the spot
diameter was bigger than the grid size, and the segments, despite having
their endpoints snapped to the grid, were smooth to a much higher precision
than the grid itself.  Despite there being a grid involved, there still were
no pixels.

Hershey also wrote about an SC~4060, which was basically the same technology
as the SC~4020 but enhanced to run at higher speed and with a wider range of
ready-made glyphs built into its aperture plate; and a "Linotron" printer,
(not clear whether it is related to the well-known Linotronic
phototypesetting machines) which operated like the SC~4020 in vector mode
but instead of the aperture plate had an optical scanner that functioned as
a read-only memory to provide raster-scanned character definitions that the
host computer could invoke at different scales.  Presumably the film in the
Linotron's optical scanner could be swapped out to provide a selection of
fonts.

The Hershey fonts that everybody remembers are the ones he suggested for use
with the SC~4060.  They consist of lists of line segments with the endpoints
on grid points.  The grid is quite coarse relative to the size of the
letters.  Hershey's guideline was 21 grid units for the height of capital
letters in large sizes; so the grid size might be thought of as comparable
to present-day screen bitmap grids.  But these vectors are neither bitmaps,
nor outlines like those in a modern "vector" font; instead they are paths
along which it is assumed the machine will sweep a circular pen.  The pen
size is only vaguely specified, but assumed to be bigger than the grid.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Pen size}

Figure~\ref{fig:capa-normal} shows a typical glyph ("occidental glyph 3001,
triplex Roman capital A") from the Hershey vector collection.  The pen size
shown is 1.59 times the grid spacing.  This seems to be close to how Hershey
intended the glyph should appear, when he originally designed it.  The wide
stroke along the side is made from three segments that overlap to form a
solidly filled wide stroke, and there are no gaps inside the filled outline
of the glyph.

\begin{figure}
  \centering\renewcommand{\SegRad}{0.79}
  \begin{tikzpicture}[scale=0.2]
    \fill[color=blue!30!white] \FatSegment{0,12}{-7,-8};
    \fill[color=blue!30!white] \FatSegment{-1,9}{5,-9};
    \fill[color=blue!30!white] \FatSegment{0,9}{6,-9};
    \fill[color=blue!30!white] \FatSegment{0,12}{7,-9};
    \fill[color=blue!30!white] \FatSegment{-5,-3}{4,-3};
    \fill[color=blue!30!white] \FatSegment{-9,-9}{-3,-9};
    \fill[color=blue!30!white] \FatSegment{2,-9}{9,-9};
    \fill[color=blue!30!white] \FatSegment{-7,-8}{-8,-9};
    \fill[color=blue!30!white] \FatSegment{-7,-8}{-5,-9};
    \fill[color=blue!30!white] \FatSegment{5,-8}{3,-9};
    \fill[color=blue!30!white] \FatSegment{5,-7}{4,-9};
    \fill[color=blue!30!white] \FatSegment{6,-7}{8,-9};
    \draw \FatSegment{0,12}{-7,-8};
    \draw \FatSegment{-1,9}{5,-9};
    \draw \FatSegment{0,9}{6,-9};
    \draw \FatSegment{0,12}{7,-9};
    \draw \FatSegment{-5,-3}{4,-3};
    \draw \FatSegment{-9,-9}{-3,-9};
    \draw \FatSegment{2,-9}{9,-9};
    \draw \FatSegment{-7,-8}{-8,-9};
    \draw \FatSegment{-7,-8}{-5,-9};
    \draw \FatSegment{5,-8}{3,-9};
    \draw \FatSegment{5,-7}{4,-9};
    \draw \FatSegment{6,-7}{8,-9};
  \end{tikzpicture}
  \caption{A typical glyph drawn with pen size 1.59 grid units.}
  \label{fig:capa-normal}
\end{figure}

Consider Figure~\ref{fig:capa-thin}, which shows a magnified view of the top
of the same glyph, drawn with a pen size of 1.26 grid units.  There is a
small, roughly triangular area uncovered inside the corner, where the
central filling stroke doesn't extend quite far enough to fill the entire
space formed by the other strokes.  The smaller the pen size, the more of
these defects occur.  However, on an actual CRT plotter, the dots don't have
sharp edges; instead of having a stark white defect in the middle of the
corner, the actual CRT plotter would produce a soft-edged light grey area
where the edges of the filling stroke partially filled in the gap.  Some of
that effect is visible in copies of Allen V.\ Hershey's original technical
report; it appears that he printed it using some kind of machine with a dot
size of somewhere around 1.2, but fuzzy edges that made these gaps less
visible.

\begin{figure}
  \centering\renewcommand{\SegRad}{0.63}
  \begin{tikzpicture}[scale=1]
    \path[clip] (-5,8) rectangle (5,13);
    \fill[color=blue!30!white] \FatSegment{0,12}{-7,-8};
    \fill[color=blue!30!white] \FatSegment{-1,9}{5,-9};
    \fill[color=blue!30!white] \FatSegment{0,9}{6,-9};
    \fill[color=blue!30!white] \FatSegment{0,12}{7,-9};
    \draw \FatSegment{0,12}{-7,-8};
    \draw \FatSegment{-1,9}{5,-9};
    \draw \FatSegment{0,9}{6,-9};
    \draw \FatSegment{0,12}{7,-9};
  \end{tikzpicture}
  \caption{Detail drawn with pen size 1.26 grid units.}
  \label{fig:capa-thin}
\end{figure}

For a modern vector font, meant for output devices with much less fuzziness,
it is not clear how to proceed.  Filling in all the gaps requires using a
pen size well over 1.6; but that will have the effect of filling in some
gaps (particularly in the more detailed Japanese kanji glyphs) that were
obviously intended not to be filled.  Reducing the pen width enough to keep
everything separated that should be, will inevitably introduce some unfilled
gaps.  There is no pen size that completely avoids both issues.  Beikaitoru
provides a selection of several different weights corresponding to different
pen sizes, but short of modifying the vectors to add extra gap-filling
strokes, no perfect solution is possible.

Many historical users of the Hershey fonts used much smaller pen sizes than
these anyway.  The Naval Weapons Lab's CRT plotters had limited resolution
and needed to plot glyphs only a few grid cells high.  Scaling the fonts was
not practical because of integer rounding concerns, anticipating the
"hinting" issues in vector fonts designed much later for use on pixel-based
displays.  In order to have different sizes of lettering, Hershey designed
complete new sets of vectors for each different size; each set would only be
used at its fixed native size.  But just a few years later, when the CAD/CAM
and GIS communities were using the Hershey fonts with pen plotters and
high-resolution vector CRTs, they had equipment capable of scaling the
fonts, and that's what they did.

It was common practice to scale the Hershey vectors to an arbitrary variable
size while not scaling the pen at all.  The pen size would usually end up
much smaller than the grid size.  This scaling would deliberately expose the
structure of the filling-in strokes that were originally meant to be hidden
by overlap.  Figure~\ref{fig:capa-vthin} shows the result.  This visual
style, much different from what appears in the original Hershey report, is
the style that many people today remember as the classic style of the
Hershey fonts.

\begin{figure}
  \centering\renewcommand{\SegRad}{0.125}
  \begin{tikzpicture}[scale=0.2]
    \fill[color=blue!30!white] \FatSegment{0,12}{-7,-8};
    \fill[color=blue!30!white] \FatSegment{-1,9}{5,-9};
    \fill[color=blue!30!white] \FatSegment{0,9}{6,-9};
    \fill[color=blue!30!white] \FatSegment{0,12}{7,-9};
    \fill[color=blue!30!white] \FatSegment{-5,-3}{4,-3};
    \fill[color=blue!30!white] \FatSegment{-9,-9}{-3,-9};
    \fill[color=blue!30!white] \FatSegment{2,-9}{9,-9};
    \fill[color=blue!30!white] \FatSegment{-7,-8}{-8,-9};
    \fill[color=blue!30!white] \FatSegment{-7,-8}{-5,-9};
    \fill[color=blue!30!white] \FatSegment{5,-8}{3,-9};
    \fill[color=blue!30!white] \FatSegment{5,-7}{4,-9};
    \fill[color=blue!30!white] \FatSegment{6,-7}{8,-9};
    \draw \FatSegment{0,12}{-7,-8};
    \draw \FatSegment{-1,9}{5,-9};
    \draw \FatSegment{0,9}{6,-9};
    \draw \FatSegment{0,12}{7,-9};
    \draw \FatSegment{-5,-3}{4,-3};
    \draw \FatSegment{-9,-9}{-3,-9};
    \draw \FatSegment{2,-9}{9,-9};
    \draw \FatSegment{-7,-8}{-8,-9};
    \draw \FatSegment{-7,-8}{-5,-9};
    \draw \FatSegment{5,-8}{3,-9};
    \draw \FatSegment{5,-7}{4,-9};
    \draw \FatSegment{6,-7}{8,-9};
  \end{tikzpicture}
  \caption{Glyph drawn with pen size 0.25 grid units.}
  \label{fig:capa-vthin}
\end{figure}

To summarize:  Hershey's report quotes dot sizes of 2.9 grid units for the
SC~4010 printer (not strictly comparable to the others because it was not a
vector printer); 2.3 for the SC~4020; and "could be as small as" 1.0 for the
SC~4060.  He also implied that smaller was better, down to the minimum of
1.0 needed to allow dots to overlap at all, which seemed not to be
achievable with 1967 technology.  It appears that Hershey's vector designs
were in fact intended for plotting with the equivalent of a sharp-edged dot
of about 1.6 grid units diameter; but in practice he would have achieved
that with a somewhat smaller fuzzy-edged dot instead.  The examples in his
report seem to have been plotted with a fuzzy dot size near 1.2, which may
have been slightly smaller than ideal.  And then subsequent to Hershey's
work, many users ended up choosing much smaller dot sizes, producing a very
different look.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Font style}

Since he was designing for hardware incapable of scaling the fonts, Hershey
designed complete alphabets at several different sizes (measured in grid
points), as well as multiple styles.  In the original designs, the sizes are
not clearly separated from the styles; and some non-alphabetic characters
seem intended for use with multiple sizes or styles.  Individual glyphs are
given four-digit numbers in a scheme unique to the Hershey report, and in
particular not corresponding to ASCII or the then-nonexistent Unicode.
function.

The main organizational division of the glyph set is into "Simplex,"
"Duplex," and "Triplex," according to how many strokes of the plotter
pen are used for the wide strokes in the glyphs.  These are then divided
into different styles, typically at least serif and sans-serif, sometimes
italic.  Most styles include upper and lowercase Latin and Greek; some also
include Cyrillic.  There are three blackletter alphabets incorrectly
described as "Gothic." There is a selection of Japanese
characters, apparently intended as a proof of concept rather than for actual
use; it is not really enough to create a complete usable font.  And there is
a collection of cartographic symbols, including (for many of the symbols)
multiple sizes and orientations, to accomodate the plotter's limitations.

The Beikaitoru fonts attempt to reorganize the glyph set according to
Unicode; each font is pieced together from multiple segments of the Hershey
set.  The styles as seen in Beikaitoru are discussed later; this summary
describes Hershey's organization.  The appendix, part, and section names and
numbering shown here are from his report.  Be warned that some of the
nomenclature does not follow modern usage.  Also, the Usenet-posted
electronic source used by Beikaitoru does not perfectly match the pictures
in the report.  The main difference is that it contains some additional
glyphs, although just a few are substituted entirely, such as number 0229,
which is lozenge in the Hershey report and centre dot in the electronic
source, while the electronic source has lozenge at 0235, a number not used
in the Hershey report.

FIXME reproduce Hershey's descriptions of his source materials.

\subsection*{Appendix B: Digitalization with vectors}

\subsubsection*{Part I: Simplex Repertory}

\paragraph*{A. FORTRAN}

Apparently a set of characters for writing FORTRAN.  Small sans-serif
uppercase-only Latin and Greek alphabets with limited punctuation; larger
upper- and lowercase Latin and Greek, upright except the lowercase Greek is
italic (as is typical in mathematics).  Large script-style upper- and
lowercase Latin.

\paragraph*{B. Cartographic}

A grab bag of line-drawing glyphs, and markers that seem to be intended for
making maps, as the title implies.  A few electronic circuit symbols.  The
electronic source also includes some specifically weather-related map
symbols not in the Hershey report.

\subsubsection*{Part II: Duplex Repertory}

\paragraph*{A. Indexical}

"Indexical" is a smaller size, probably intended for subscripts and
superscripts.  Upper and lowercase Latin and Greek (upright except the
lowercase Greek is italic); upper and lowercase italic Latin.  All are
serif, and seem to be based on Scotch Modern.  Some additional typographic
and astronomical symbols.

\paragraph*{B. Principal}

The same styles as for "indexical," larger.  Some additional math and music
symbols.  Oversize math brackets, sum, product, root, integral.  Sans serif
large upright Latin; script Latin.  Serif upright Cyrillic.

\subsubsection*{Part III: Triplex Repertory}

Three styles of Latin blackletter (not "Gothic" despite the titles), upper-
and lowercase.  The Hershey report shows only the alphabets, but the
electronic source, as well as numbering them differently, also
includes a single set of blackletter numerals and punctuation
apparently intended to be shared by all three styles.

\paragraph*{A. English Gothic}

\paragraph*{B. Italian Gothic}

\paragraph*{C. German Gothic}

\subsection*{Appendix C: Digitalization of Japanese}

\subsubsection*{Part I: Kanji List}

A selection of 603 kanji characters, apparently intended as a representative
sampling for proof of concept, determined as follows: those of the 214
traditional radicals that are characters in themselves and included on the
Touyou Kanji list (i.e.\ the Japanese government's "daily use" list current
at the time; the modern Jouyou Kanji were introduced in 1981); the
first-grade school kanji (again, current as of 1967); and "a selection of
characters of scientific interest." Apparently some effort was made to
include both halves of pairs of antonyms if they chose to include either
half, and any components shared by more than one character they would
otherwise include; the details of these rules were not elaborated in the
report.  The report and the electronic source index these characters by
Nelson dictionary numbers.

This range also includes three punctuation marks:  period and comma, both of
which are Western-style shapes but spaced for use with Japanese, and kana
vowel length mark.

\subsubsection*{Part II: Hiragana List}

The full basic set of hiragana, including archaic glyphs for "wi" and "we"
and an extra copy of the "u" glyph in a spot where the character code
implies "wu"; separate glyphs with dakuten and handakuten where applicable. 
No "small" kana.  

\subsubsection*{Part III: Katakana List}

Katakana, covering the same range as the hiragana.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Guide to the Beikaitoru fonts}

Reviving the Hershey fonts for use with 21st Century technology required
rearranging and coding the glyphs to fit in the framework of Unicode,
splitting the "repertories" into distinct font styles, and making decisions
on issues like dot size that in the 1960s would have been dictated by
hardware considerations.  Making many of these choices available to users
implies the creation of a large family of fonts, and then there is some
question of how to name and identify the different fonts within the family.

The fonts created by this package are all named "Beikaitoru" with a
three-digit decimal number identifying the specific font.  These numbers are
defined as sums of integers encoding information about the font:
\begin{itemize}
   \item a multiple of 50 (from 100 to 800) defining the general style;
   \item plus 10 for coordinate smoothing;
   \item plus a one-digit number (from 0 to 9) defining the dot or pen size.
\end{itemize}

The multiples of 50 encoding general style are selected from the following
list.  The style definitions are work in progress.  As of this writing, only
those marked with an asterisk are defined and will build correctly; others
may build with placeholder code copied from other styles (i.e.\ the glyphs
in the font are not correctly the style described by the number), or they
may not build at all.  Even for styles that build completely enough to be
usable, not all characters planned for eventual inclusion will necessarily
be included at the present time.

\begin{tabular}{ll}
  code & description \\ \hline
  100* & simplex small grid \\
  150* & simplex medium grid \\
  200* & simplex script \\
  250 & [not used] \\
  300* & duplex roman indexical (small grid) \\
  350 & duplex italic indexical (small grid) \\
  400* & duplex roman principal (large grid) \\
  450 & duplex italic principal (large grid) \\
  500* & duplex sans serif \\
  550 & duplex script \\
  600* & triplex roman \\
  650 & triplex italic \\
  700 & triplex German blackletter \\
  750 & triplex English blackletter \\
  800 & triplex Italian blackletter
\end{tabular}

Since these are scaleable vector fonts, the actual sizes of text will be
determined by the scaling chosen; references to size above refer to the
sizes of the glyphs in terms of Hershey's grid units.  Fonts defined on a
smaller grid will appear bolder and blockier than fonts on a larger grid, at
the same physical size.  These fonts do not directly support any concept of
"optical size," although using smaller grid sizes for smaller optical sizes
will produce a similar effect.

Adding 10 for curve smoothing invokes a special feature of Beikaitoru's
Metapost code that attempts to detect when the Hershey plotter data is
approximating a curve with many small line segments, and in such cases
replaces the line segments with interpolating cubic splines.  In many
glyphs, the results look more visually appealing.  However, the code
sometimes guesses wrong, and the jagged effect of the original line segments
may be desirable to impart an historical flavour anyway, so this feature is
not necessarily always a good thing.

Beikaitoru provides the following nine numbered sizes, each of which defines
a diameter for the dot or pen in grid units.  The last digit of
each font's style number is the pen size; for instance, Beikaitoru 406 uses
pen size 6.

\begin{tabular}{ll}
  size & dot diameter \\ \hline
  1 & 0.125 \\
  2 & 0.250 \\
  3 & 0.500 \\
  4 & 1.000 \\
  5 & 1.260 \\
  6 & 1.587 \\
  7 & 2.000 \\
  8 & 2.520 \\
  9 & 3.175
\end{tabular}

Size 6 is probably optimal for simulating the way Hershey intended his fonts
to look on the hardware that was current in 1967, and it is the size used to
typeset the main text of this documentation.  (This text is set in
Beikaitoru~406.)  Sizes 1, 2, and 3, depending
on scaling, may be appropriate for simulating later historical uses of the
Hershey vectors in vector-plotting environments.

\end{document}
